\chapter{Conclusion}
\label{chap:conclusion}
%Efficient and productive functional programming requires efficient and
%expressive functional data structures.
Programming languages require support for efficient data structures. In
particular, the data structures should support programming in the
paradigm that the language supports. Racket and its statically typed
dialect Typed Racket are functional languages, but lack many functional
data structures. In this thesis, I have described a comprehensive
library of efficient functional data structures for Typed Racket and an
evaluation of Typed Racket as a language for developing large libraries
of typed programs.

\section{Contributions}
In support of this MS thesis, I implementated a comprehensive library of
data structures in Typed Racket. Chapter \ref{chap:four} shows that the
functional data structures are as efficient as their imperative
counterparts, supporting efficient programming in the functional
paradigm.

The evaluation of Typed Racket shows that even though Typed Racket has a
type system built to support the Racket idioms, it supports
straightforward porting of code from other statically typed functional
languages with different type systems, such as Haskell and
ML. Additionally Typed Racket is easy to learn and use.

%hope that this enables programmers to write functional programs, and
%inspires library writers to use functional designs and to produce new
%libraries to enable functional programming.
\section{Future Work}
As mentioned earlier, because Typed Racket does not support polymorphic
recursion, it is not possible to implement some data structures, such as
finger trees \citep{finger}. Support for polymorphic recursion will be
useful. It may also improve the performance of some of the existing data
structures.

The library of data structures developed is a comprehensive foundation
for implementing data structures that support parallel programming. It
remains to be seen how these data structures perform in Typed Racket and
much more work remains to be done in evaluating Typed Racket in the
context of these data structures that support parallel programming.

